---
title: "R Notebook"
output: html_notebook
---



```{r}
library(plyr)
library(coin)

```


```{r}
# Read in data
data <- read.table("bptrial.txt", header=TRUE, sep=",", dec=".")
original <- data
# Set column names
colnames(data) <- c("treatment", "sex", "weight", "age",
                    "comp", "dbpdif", "dbp3", "dbp2", "dbp1")
# Add column dbp_end for the 5th measurement at the end of the experiment
data$dbp_end <- data$dbp3 + data$dbpdif
# Mean dbp during the run-in period
data$dbp_mean <- round((data$dbp1 + data$dbp2 + data$dbp3) / 3, 1)
# Raw data subset
data_raw <- data
# Treatment group as factor
data$treatment <- as.factor(mapvalues(
  data$treatment,
  from=c(1, 2, 3),
  to=c("Treatment 1", "Placebo", "Treatment 3")
))
# Gender as factor
data$sex <- as.factor(mapvalues(data$sex, from=c(0, 1), to=c("Female", "Male")))
# 80% adherence rule
data <- data[data$comp >= 0.8,]
# Subset of treatment 1 and the placebo
data12 <- data[data$treatment != "Treatment 3",]
levels(data12$treatment) <- c("Treatment 1", "Placebo", NA)
```

## 2. 

```{r}
# Amount of possible permutations
choose(84, 42)
```

```{r}
# Get observed
t.obs_dif <- t.test(data12$dbpdif~data12$treatment)$statistic
t.obs_end <- t.test(data12$dbp_end~data12$treatment)$statistic
t.obs_dif
t.obs_end
```

# Parametric

We probably don't need to do this one

```{r}
N <- 10000
t.approx_dif <- rep(NA,N)
t.approx_end <- rep(NA,N)
perm.data <- data12
for(i in 1:N) {
  # Set seed for reproducibility
  set.seed(1991 + i)
  # Randomly assign treatment group
  perm.data$group <- sample(perm.data$treatment)
  # T-test for dbpdif and dbp5
  test_dif <- t.test(dbpdif~group, data=perm.data)
  test_end <- t.test(dbp_end~group, data=perm.data)
  # Store T value
  t.approx_dif[i]<-test_dif$statistic
  t.approx_end[i]<-test_end$statistic
}
```

```{r}
mean(abs(t.approx_dif) >= abs(t.obs_dif))
mean(abs(t.approx_end) >= abs(t.obs_end))
```


# Non-parametric

We probably have to change it to one-sided?

## Wilcoxon

Asymptotic null distribution

```{r}
wilcox.test(dbpdif~treatment, data=data12, exact=FALSE, correct=FALSE)
wilcox.test(dbp_end~treatment, data=data12, exact=FALSE, correct=FALSE)
```

The p-value is smaller when using the final DBP measure opposed to using the difference between DBP measures 3 and 5.

## Manual

Different result?

```{r}
# Rank data
data12$rank_dif <- rank(data12$dbpdif)
data12$rank_end <- rank(data12$dbp_end)
# Sum of ranks
w_dif <- sum(data12$rank_dif[data12$treatment == "Treatment 1"])
w_end <- sum(data12$rank_end[data12$treatment == "Treatment 1"])
# Amount of observations
n_treatment <- table(data12$treatment)["Treatment 1"]
n_placebo <- table(data12$treatment)["Placebo"]
n <- n_treatment + n_placebo
exp <- (n_treatment * (n + 1)) / 2
var <- (n_treatment * n_placebo * (n + 1)) / 12
standardized_dif <- (w_dif - exp) / sqrt(var)
standardized_end <- (w_end - exp) / sqrt(var)
standardized_dif
standardized_end
```

```{r}
criticalValues <- c(qnorm(0.025), qnorm(0.975))
criticalValues
```

```{r}
2 * (1 - pnorm(abs(standardized_dif)))
2 * (1 - pnorm(abs(standardized_end)))
```

The p-value is smaller when using the final DBP measure opposed to using the difference between DBP measures 3 and 5.

## Coin package

```{r}
# no
wilcox_test(data12$dbpdif~data12$treatment, distribution="exact")
wilcox_test(data12$dbp_end~data12$treatment, distribution="exact")
# no
wilcox_test(data12$dbpdif~data12$treatment, distribution=approximate(B = 10000000))
wilcox_test(data12$dbp_end~data12$treatment, distribution=approximate(B = 10000000))
# yes
# Outcome is equal to using the built-in wilcoxon formula
wilcox_test(data12$dbpdif~data12$treatment, distribution="asymptotic")
wilcox_test(data12$dbp_end~data12$treatment, distribution="asymptotic")
```

